---
title: "NHL API Vignette"
author: "John Clements"
date: "6/15/2021"
output: 
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Requirements

To use the functions for interacting with the [NHL API](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md), you need to install and load:

- [`tidyverse`](https://www.tidyverse.org/): tons of useful features for data manipulation and plotting
- [`jsonlite`](https://cran.r-project.org/web/packages/jsonlite/): API interaction

Additionally, I used the following packages for the data exploration:

- [`cowplot`](https://cran.r-project.org/web/packages/cowplot/index.html): extra functionality for `ggplot2`
- [`imager`](https://cran.r-project.org/web/packages/imager/): loading in an images
- [`knitr`](https://cran.r-project.org/web/packages/knitr/index.html)

I didn't pull in the entire  to avoid pulling in unnecessary functions.

```{r warning=FALSE, message=FALSE}
library(jsonlite)
library(tidyverse)
library(cowplot)
library(imager)
```

# Functions

## `convertToNumeric`

I made this helper function to convert columns that contain numeric data stored as `character` values to numeric data types. I ran into an issue where my calls from the API were returning some numeric data as `character` data and needed a way to deal with this issue without just calling `as.numeric` as needed.

```{r}
convertToNumeric <- function(vec){
  ###
  # This function will convert the input vector to a numeric vector if it is 
  # able to. Otherwise, it just returns the vector.
  ###
  
  # If any of the values in vec return NA when trying to convert to numeric,
  # just return vec as the output.
  if (any(is.na(suppressWarnings(as.numeric(vec))) == TRUE)){
    output <- vec
  }
  # Otherwise, convert vec to a numeric vector.
  else {
    output <- as.numeric(vec)
  }
  return(output)
}
```

## `franchise`

I wrote this function to interact with the `franchise` endpoint of the [NHL Records API](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md). It returns a `data.frame` containing the franchise and current team Id numbers, the name and abbreviations of teams, and their first and last seasons. It takes one argument; `team`, which can either be `"all"`, the full name of a team (e.g. `"New Jersey Devils"`), or the most recent team Id (e.g. `1` for the New Jersey Devils).

```{r}
franchise <- function(team="all"){
  ###
  # This functions returns a dataframe with metadata on all teams who ever 
  # played in the NHL. It can also return those columns for a single team if a 
  # franchise ID or name is passed.
  ###
  
  # Get the franchise data from the franchises endpoint.
  outputAPI <- fromJSON(
      "https://records.nhl.com/site/api/franchise"
      )
  
  # Select only the dataframe.
  output <- outputAPI$data
  
  # If team does not equal "all", check if it is a franchise ID or team name.
  if (team != "all"){
    
    # If team is in the id column, subset output for just that row.
    if (team %in% output$id){
      output <- output %>%
        filter(mostRecentTeamId == team)
    }
    # If team is in the fullName column, subset output for just that row.
    else if (team %in% output$fullName){
      output <- output %>%
        filter(fullName == team)
    }
    # Otherwise, warn the user and return the entire dataframe.
    else {
      message <- paste("ERROR: Argument for team was not found in either",
                       "the fullName or id columns. Try franchise('all') to",
                       "find the team you're looking for.")
      stop(message)
    }
  }
  # Do nothing if the team value equals "all".
  else {
    
  }
  
  # Convert any columns that should be numeric to numeric.
  output <- suppressMessages(as.data.frame(lapply(output, convertToNumeric)))
  
  # Return the output dataframe.
  return(output)
}
```

## `teamTotals`

I wrote this function to interact with the `franchise-team-totals` endpoint of the NHL Records API. It returns a large number of statistics for the entire history of a team for both regular season and playoff games. It takes one argument; `team`, which can either be `"all"`, the full name of a team (e.g. `"New Jersey Devils"`), or the most recent team Id (e.g. `1` for the New Jersey Devils).

```{r}
teamTotals <- function(team="all"){
  ###
  # This function returns total stats for every franchise (ex roadTies,
  # roadWins, etc) unless a specific team Id or full team name is passed.
  ###
  
  # Get the franchise data from the franchises endpoint.
  outputAPI <- fromJSON(
    "https://records.nhl.com/site/api/franchise-team-totals"
    )
  
  # Select only the dataframe.
  output <- outputAPI$data
  
  # If team does not equal "all", check if it is a team ID or team name.
  if (team != "all"){
    
    # If team is in the franchiseId column, subset output for just that row.
    if (team %in% output$teamId){
      output <- output %>%
        filter(teamId == team)
    }
    # If team is in the teamName column, subset output for just that row.
    else if (team %in% output$teamName){
      output <- output %>%
        filter(teamName == team)
    }
    # Otherwise, warn the user and return the entire dataframe.
    else {
      message <- paste("WARNING: Argument for team was not found in either",
                       "the teamName or franchiseId columns. Returning all",
                       "franchises.")
      warning(message)
    }
  }
  # Do nothing if the team value equals "all".
  else {
    
  }
  
  # Convert any columns that should be numeric to numeric.
  output <- suppressMessages(as.data.frame(lapply(output, convertToNumeric)))
  
  # Return the output dataframe.
  return(output)
}
```

## `findTeamId`

This is a helper function to look up the most recent team Id for a full team name (e.g. `"Boston Bruins"`) because I a didn't feel like hard coding it. It is used in the `franchiseDetail` and `seasonStats` functions because the endpoints they connect to use the most recent team Id to access team-level data in the API call.

```{r}
findTeamId <- function(teamName){
  ###
  # This helper function looks up the franchise id for a full team name 
  # e.g. "Boston Bruins".
  ###
  
  # Call the franchise function with the team name.
  teamId <- franchise(teamName)
  
  # Return only the most recent team id number.
  return(teamId$mostRecentTeamId)
}
```

## `findFranchiseId`

This is a helper function does the same thing as `findTeamId`, except for the franchise Id. It is used in the `seasonRecords`, `goalieRecords`, and `skaterRecords` functions because they use the franchise Id to access team-level data in the API call.

```{r}
findFranchiseId <- function(teamName){
  ###
  # This helper function looks up the franchise id for a full team name 
  # e.g. "Boston Bruins".
  ###
  
  # Call the franchise function with the team name.
  teamId <- franchise(teamName)
  
  # Return only the most recent team id number.
  return(teamId$id)
}
```

## `seasonRecords`

The `seasonRecords` function returns the records stats for a single team. For example, the most goals scored in a season and the season they scored it in. It takes one argument; `team`, which can either be the full name of a team (e.g. `"New Jersey Devils"`), or the **franchise Id** (e.g. `23` for the New Jersey Devils).

```{r}
seasonRecords <- function(team){
  ###
  # This functions returns a dataframe with the season records for a variety 
  # of stats for a single team.
  ###
  
  # If team is a "character" type, try to look up the team id.
  if (typeof(team) == "character"){
    teamId = findFranchiseId(team)
  }
  # If team is an integer, set teamId equal to team.
  else if ((typeof(team) == "double") & (team %% 1 == 0)){
    teamId = team
  }
  # Otherwise, throw an error.
  else {
    message <- paste("Please pass a team id (integer) or a full team name",
                     "e.g. 'Boston Bruins'.")
    stop(message)
  }
  
  # Set the base url, endpoint, and combine them with teamId for the full url.
  baseURL <- "https://records.nhl.com/site/api/"
  endpoint <- "franchise-season-records?cayenneExp=franchiseId="
  fullURL <- paste0(baseURL, endpoint, teamId)
  
  # Get the API output.
  outputAPI <- fromJSON(fullURL)
  # Select only the data.
  output <- outputAPI$data
  
  # Convert any columns that should be numeric to numeric.
  output <- suppressMessages(as.data.frame(lapply(output, convertToNumeric)))
  
  # Return the output from the request.
  return(output)
}
```

## `goalieRecords`

The `goalieRecords` function returns the stats for all the goalies who've played for a single team. For example, the most saves in a game and the date of the game. It takes one argument; `team`, which can either be the full name of a team (e.g. `"New Jersey Devils"`), or the **franchise Id** (e.g. `23` for the New Jersey Devils).

```{r}
goalieRecords <- function(team){
  ###
  # This functions returns a dataframe with the goalie records for a team.
  ###
  
  # If team is a "character" type, try to look up the team id.
  if (typeof(team) == "character"){
    teamId = findFranchiseId(team)
  }
  # If team is an integer, set teamId equal to team.
  else if ((typeof(team) == "double") & (team %% 1 == 0)){
    teamId = team
  }
  # Otherwise, throw an error.
  else {
    message <- paste("Please pass a team id (integer) or a full team name",
                     "e.g. 'Boston Bruins'.")
    stop(message)
  }
  
  # Set the base url, endpoint, and combine them with teamId for the full url.
  baseURL <- "https://records.nhl.com/site/api/"
  endpoint <- "franchise-goalie-records?cayenneExp=franchiseId="
  fullURL <- paste0(baseURL, endpoint, teamId)
  
  # Get the API output.
  outputAPI <- fromJSON(fullURL)
  # Select only the data.
  output <- outputAPI$data
  
  # Convert any columns that should be numeric to numeric.
  output <- suppressMessages(as.data.frame(lapply(output, convertToNumeric)))
  
  # Return the output from the request.
  return(output)
}
```

## `skaterRecords`

The `skaterRecords` function returns the stats for all the non-goalie players who've played for a single team. For example, the most penalty minutes they received in one season and the season the season it occurred in. It takes one argument; `team`, which can either be the full name of a team (e.g. `"New Jersey Devils"`), or the **franchise Id** (e.g. `23` for the New Jersey Devils).

```{r}
skaterRecords <- function(team){
  ###
  # This functions returns a dataframe with the skater records for a team.
  ###
  
  # If team is a "character" type, try to look up the team id.
  if (typeof(team) == "character"){
    teamId = findFranchiseId(team)
  }
  # If team is an integer, set teamId equal to team.
  else if ((typeof(team) == "double") & (team %% 1 == 0)){
    teamId = team
  }
  # Otherwise, throw an error.
  else {
    message <- paste("Please pass a team id (integer) or a full team name",
                     "e.g. 'Boston Bruins'.")
    stop(message)
  }
  
  # Set the base url, endpoint, and combine them with teamId for the full url.
  baseURL <- "https://records.nhl.com/site/api/"
  endpoint <- "franchise-skater-records?cayenneExp=franchiseId="
  fullURL <- paste0(baseURL, endpoint, teamId)
  
  # Get the API output.
  outputAPI <- fromJSON(fullURL)
  # Select only the data.
  output <- outputAPI$data
  
  # Convert any columns that should be numeric to numeric.
  output <- suppressMessages(as.data.frame(lapply(output, convertToNumeric)))
  
  # Return the output from the request.
  return(output)
}
```

## `franchiseDetail`

This function gets info for a franchise, like their retired numbers. It takes one argument; `team`, which can either be the full name of a team (e.g. `"New Jersey Devils"`), or the most recent team Id (e.g. `1` for the New Jersey Devils).

```{r}
franchiseDetail <- function(team){
  ###
  # This functions returns a dataframe with the data for a team.
  ###
  
  # If team is a "character" type, try to look up the team id.
  if (typeof(team) == "character"){
    teamId = findTeamId(team)
  }
  # If team is an integer, set teamId equal to team.
  else if ((typeof(team) == "double") & (team %% 1 == 0)){
    teamId = team
  }
  # Otherwise, throw an error.
  else {
    message <- paste("Please pass a team id (integer) or a full team name",
                     "e.g. 'Boston Bruins'.")
    stop(message)
  }
  
  # Set the base url, endpoint, and combine them with teamId for the full url.
  baseURL <- "https://records.nhl.com/site/api/"
  endpoint <- "franchise-detail?cayenneExp=mostRecentTeamId="
  fullURL <- paste0(baseURL, endpoint, teamId)
  
  # Get the API output.
  outputAPI <- fromJSON(fullURL)
  # Select only the data.
  output <- outputAPI$data
  
  # Convert any columns that should be numeric to numeric.
  output <- suppressMessages(as.data.frame(lapply(output, convertToNumeric)))
  
  # Return the output from the request.
  return(output)
}
```

## `seasonStats`

I wrote this function to interact with the `franchise` endpoint of the [NHL Stats API](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md). It returns a `data.frame` containing the current season stats. It takes one argument; `team`, which can either be `"all"`, the full name of a team (e.g. `"New Jersey Devils"`), or the most recent team Id (e.g. `1` for the New Jersey Devils).

```{r}
seasonStats <- function(team="all", raw=FALSE){
  ###
  # Returns the current seasons stats for all teams or just one. If raw is 
  # FALSE, it returns only the stats. If raw is TRUE, it just returns the API
  # output for the user to parse.
  ###
  
  # If raw is not a valid input, throw an error.
  if ((raw != TRUE) & (raw != FALSE)){
    stop("raw must equal TRUE or FALSE!")
  }
  # Otherwise do nothing.
  else {
    
  }
  
  # If team equals "all" the spot where ID goes in the URL will be blank.
  if (team == "all"){
    teamId = ""
  }
  # If team is a "character" type, try to look up the team id.
  else if (typeof(team) == "character"){
    teamId = findTeamId(team)
  }
  # If team is an integer, set teamId equal to team.
  else if ((typeof(team) == "double") & (team %% 1 == 0)){
    teamId = team
  }
  # Otherwise, throw an error.
  else {
    message <- paste("Please pass a team id (integer) or a full team name",
                     "e.g. 'Boston Bruins'.")
    stop(message)
  }
  
  # Paste together the URL.
  baseURL <- "https://statsapi.web.nhl.com/api/v1/teams/"
  modifier <- "?expand=team.stats"
  fullURL <- paste0(baseURL, teamId, modifier)
  # Get the data from the endpoint.
  outputAPI <- fromJSON(fullURL, flatten=TRUE)
  
  if (team != "all"){
    
    # If raw is FALSE, give back only the stats.
    if (raw == FALSE){
      teamStats <- outputAPI$teams$teamStats[[1]]$splits[[1]][1,]
      # Convert any columns that should be numeric to numeric.
      teamStats <- suppressMessages(as.data.frame(lapply(teamStats,
                                                         convertToNumeric)))
    }
    # If the user wants the raw API output, give it to them.
    else {
      teamStats <- outputAPI$teams
    }
  }
  else {
    
    # If raw is FALSE, give back only the stats.
    if (raw == FALSE){
      # Get the teamStats list where each element is a data.frame.
      output <- outputAPI$teams$teamStats
      # Count the number of teams. The last element is NULL.
      num_teams = length(output) - 1
      # Make a variable to hold just the stats, starting with the first team.
      teamStats <- output[[1]]$splits[[1]][1,]
      
      # Loop through the 2nd to the last team in the list.
      for (i in seq(2, num_teams)){
        # Select only the first row of the seasons stats for the team.
        stats <- output[[i]]$splits[[1]][1,]
        # Add the row to teamStats.
        teamStats <- rbind(teamStats, stats)
      }
    # Convert any columns that should be numeric to numeric.
    teamStats <- suppressMessages(as.data.frame(lapply(teamStats,
                                                       convertToNumeric)))
    }
    # If the user wants the raw API output, give it to them.
    else {
      teamStats <- outputAPI$teams
    }
  }
  
  # Return team stats.
  return(teamStats)
}
```

## `nhlAPI`

This function is a wrapper function for all the others above. You simply pass the name of the function you want to use, like `"seasonStats`, and any additional arguments for that function.

```{r}
nhlAPI <- function(func, ...){
  ###
  # This function is a wrapper for the other functions. It takes in the name
  # of the function to use as a character and any additional arguments for that
  # function.
  ###
  
  # Find and call the appropriate function.
  
  if (func == "franchise"){
    output <- franchise(...)
  }
  else if (func == "teamTotals"){
    output <- teamTotals(...)
  }
  else if (func == "seasonRecords"){
    output <- seasonRecords(...)
  }
  else if (func == "goalieRecords"){
    output <- goalieRecords(...)
  }
  else if (func == "skaterRecords"){
    output <- skaterRecords(...)
  }
  else if (func == "franchiseDetail"){
    output <- franchiseDetail(...)
  }
  else if (func == "seasonStats"){
    output <- seasonStats(...)
  }
  else {
    stop("Argument for func is not valid!")
  }
  return(output)
}
```

# Data Exploration

Now that we can interact with a few of the endpoints of the NHL API, let's get some data from them.

First, let's pull current season stats for all teams by calling `nhlAPI("seasonStats")`.

```{r }
# Get the current season stats for all of the teams in the NHL.
currentSeason <- nhlAPI("seasonStats")
```

Two variables of interest to me are shots per game and shooting percentage. I'm interested in how these two stats relate to a team's win-loss ratio. This variable doesn't exist, so I need to calculate it. The formula is $\text{win-loss-ratio} = \frac{wins}{losses}$. A win-loss-ratio of 1 means a team has an even record. Less than one is a losing record and more than 1 is a winning record.

```{r}
# Add a column for the win/loss ratio (wins per loss)
currentSeason <- currentSeason %>%
  mutate(win.loss.ratio = stat.wins / stat.losses)
```

My guess is these two are positively related to the win-to-loss ratio. A high shooting percentage means you are fairly likely to score when you shoot. A high number of shots per game likely means you are controlling the puck more than the other team and you have more scoring opportunities, unless you're taking wild shots. Plus, I think there is a quote from some hockey player about taking shots.

```{r}
# Read in the image.
img <- load.image("./images/Wayne Gretzky Quote.jpg")
# Display the image.
plot(img, axes=0)
```

Below I plotted the win-loss ratio against shots per game and shooting percentage. I added a regression line as well. As expected, both are positively related to the win-loss-ratio, although there appears to be some [heteroskedasticity](https://en.wikipedia.org/wiki/Heteroscedasticity) in the plot on the right.

```{r}
# Create a ggplot object to plot shooting pct vs. shots per game.
plot1 <- ggplot(currentSeason, 
               aes(stat.shotsPerGame,
                   win.loss.ratio,
                   color=win.loss.ratio))
# Make a scatter plot.
plot1 <- plot1 + geom_point(size=4, alpha=0.75)
# Adjust the color gradient to go from blue at the low end, to red at the high
# end of win.loss.ratio
plot1 <- plot1 + scale_color_gradient(low="blue", high="red")
# Remove the legend.
plot1 <- plot1 + theme(legend.position="none")
# Add a regression line.
plot1 <- plot1 + geom_smooth(method=lm, formula=y~x, color="black")
# Add an x-axis label.
plot1 <- plot1 + scale_x_continuous("Shots per Game")
# Add an y-axis label.
plot1 <- plot1 + scale_y_continuous("Wins-Loss Ratio")
# Add a title.
plot1 <- plot1 + ggtitle("Wins/Losses vs. Shots per Game")

# Create a ggplot object to plot shooting pct vs. shots per game.
plot2 <- ggplot(currentSeason, 
               aes(stat.shootingPctg,
                   win.loss.ratio,
                   color=win.loss.ratio))
# Make a scatter plot.
plot2 <- plot2 + geom_point(size=4, alpha=0.75)
# Adjust the color gradient to go from blue at the low end, to red at the high
# end of win.loss.ratio
plot2 <- plot2 + scale_color_gradient(low="blue", high="red")
# Remove the legend.
plot2 <- plot2 + theme(legend.position="none")
# Add a regression line.
plot2 <- plot2 + geom_smooth(method=lm, formula=y~x, color="black")
# Add an x-axis label.
plot2 <- plot2 + scale_x_continuous("Shooting Percentage")
# Add an y-axis label.
plot2 <- plot2 + scale_y_continuous("Wins-Loss Ratio")
# Add a title.
plot2 <- plot2 + ggtitle("Wins/Losses vs. Shooting Perc.")

# Plot them both together.
plot_grid(plot1, plot2, ncol=2)
```

Now let's look at shooting percentage vs. shots per game. I added a color gradient for the win-loss-ratio. There doesn't appear to be a clear relationship between those two variables.

```{r}
# Create a ggplot object to plot shooting pct vs. shots per game.
plot3 <- ggplot(currentSeason, 
               aes(stat.shotsPerGame,
                   stat.shootingPctg,
                   color=win.loss.ratio))
# Make a scatter plot.
plot3 <- plot3 + geom_point(size=4, alpha=0.75)
# Adjust the color gradient to go from blue at the low end, to red at the high
# end of win.loss.ratio
plot3 <- plot3 + scale_color_gradient(low="blue", high="red",
                                    name="Wins / Losses")
# Add an x-axis label.
plot3 <- plot3 + scale_x_continuous("Shots per Game")
# Add an y-axis label.
plot3 <- plot3 + scale_y_continuous("Shooting Percentage")
# Add a title.
plot3 <- plot3 + ggtitle("Shooting Percentage vs. Shots per Game")

plot3
```

Now let's look at some total season stats. I called `nhlAPI("teamTotals")` to get this data.

```{r}
# Get some stats for the total history of a team.
teamTotalStats <- nhlAPI("teamTotals")
```

First, let's compare the number of inactive vs. active teams. There are 26 inactive and 31 active teams. Almost as many teams have left the NHL than are currently still playing.

```{r}
table(teamTotalStats %>%
  filter(gameTypeId == 2) %>%
  mutate(currentlyActive = ifelse(is.na(lastSeasonId), 
                                  "Active", "Inactive")) %>%
  select(currentlyActive))
```
Not all the teams have been around for the same length of time and I want to adjust some stats, say total penalty minutes, to be able to compare teams on the same basis. 

To do this, I'm calculating the the penalty minutes per game as $\frac{\text{Total Penalty Minutes}{Total Games Played}}$. I again calculate the win-loss ratio, because I want to see how that is related to penalty minutes per game. I'm also interested in how a game being a playoff or regular season game affects penalty minutes. I think the importance of a playoff game might cause changes in the penalty behavior.

```{r}
teamTotalStats <- teamTotalStats %>%
  # Calculate the penalty minutes per game.
  mutate(penaltyMinutesPerGame = penaltyMinutes / gamesPlayed,
         winsLossRatio = wins / losses,
         gameType = ifelse(gameTypeId == 2, "Regular Season", "Playoffs"),
         pointsPerGame = points / gamesPlayed)
```

```{r}
pointsSumm <- teamTotalStats %>% 
  select(gameType, pointsPerGame) %>%
  filter(is.na(pointsPerGame) == FALSE) %>%
  group_by(gameType) %>%
  summarize("Min." = min(pointsPerGame),
            "1st Quartile" = quantile(pointsPerGame, 0.25, na.rm=TRUE),
            "Median" = quantile(pointsPerGame, 0.5, na.rm=TRUE),
            "Mean" = mean(pointsPerGame, na.rm=TRUE),
            "3rd Quartile" = quantile(pointsPerGame, 0.75, na.rm=TRUE),
            "Max" = max(pointsPerGame),
            "Std. Dev." = sd(pointsPerGame, na.rm=TRUE)
            )

knitr::kable(pointsSumm, 
             caption="Summary Statistics for Penalty Minutes per Game by Game Type",
             digits=2)
```

Now let's look at a numeric summary of per game penalty minutes by game type. Playoff games seem to have a higher central tendency and be more variable in the penalty minutes per game.

```{r}
penMinSumm <- teamTotalStats %>% 
  select(gameType, penaltyMinutesPerGame) %>%
  group_by(gameType) %>%
  summarize("Min." = min(penaltyMinutesPerGame),
            "1st Quartile" = quantile(penaltyMinutesPerGame, 0.25),
            "Median" = quantile(penaltyMinutesPerGame, 0.5),
            "Mean" = mean(penaltyMinutesPerGame),
            "3rd Quartile" = quantile(penaltyMinutesPerGame, 0.75),
            "Max" = max(penaltyMinutesPerGame),
            "Std. Dev." = sd(penaltyMinutesPerGame)
            )

knitr::kable(penMinSumm, 
             caption="Summary Statistics for Penalty Minutes per Game by Game Type",
             digits=2)
```

The box plot below provides the same information as in the table above, minus the mean and standard deviation. It's clear playoff games tend to have higher penalty minutes per game and the right tail stretches decently far.

```{r}
plot4 <- ggplot(teamTotalStats, 
               aes(gameType,
                   penaltyMinutesPerGame,
                   color=gameType))
plot4 <- plot4 + geom_boxplot() + geom_jitter()
plot4 <- plot4 + scale_x_discrete("Game Type")
plot4 <- plot4 + scale_y_continuous("Penalty Minutes per Game")
plot4 <- plot4 + ggtitle("Penalty Minutes per Game by Game Type")
plot4 <- plot4 + theme(legend.position = "none")

plot4
```
The histogram below provides another way to visualize the distributions of penalty minutes per game.

```{r}
plot5 <- ggplot(teamTotalStats,aes(penaltyMinutesPerGame, fill=gameType)) + 
    geom_histogram(data=subset(teamTotalStats,gameType == 'Regular Season'),
                   bins=10, alpha = 0.5) +
    geom_histogram(data=subset(teamTotalStats,gameType == 'Playoffs'),
                   bins=10, alpha = 0.5) + 
   guides(fill=guide_legend(title="Game Type")) +
    scale_x_continuous("Penalty Minutes per Game") + 
    scale_y_continuous("Count") +
    ggtitle("Histogram of Penalty Minutes per Game by Game Type")

plot5
```

I'm curious as to who are the teams spending the most time in the penalty box per game. I filtered for active teams and regular season games and made a bar chart for each time and their penalty minutes per game. I sorted the bars in the bar chart from worst behaved to best behaved.

```{r}
mostPenaltyMinsRegSeason <- teamTotalStats %>%
  # Filter for active teams and regular season games.
  filter(is.na(lastSeasonId) & (gameTypeId == 2)) %>%
  # Sort from most penalty minutes per game to the least.
  arrange(desc(penaltyMinutesPerGame)) %>%
  # select the triCode column.
  select(triCode)

plot6 <- teamTotalStats %>%
  filter(is.na(lastSeasonId) & (gameTypeId == 2)) %>%
  mutate(sortedTriCode  = factor(triCode,
                                 levels=mostPenaltyMinsRegSeason[["triCode"]],
                                    ordered=TRUE))  %>%
  ggplot(aes(sortedTriCode, penaltyMinutesPerGame, 
             fill=penaltyMinutesPerGame)) +
  geom_col() + 
  theme(axis.text.x=element_text(angle=90), legend.position="none") +
  scale_fill_gradient(low="blue", high="red") +
  scale_x_discrete("Team") + 
  scale_y_continuous("Penalty Minutes per Game") +
  ggtitle("Penalty Minutes per Regular Season Game by Team")

plot7 <- teamTotalStats %>%
  filter(is.na(lastSeasonId) & (gameTypeId == 3)) %>%
  mutate(sortedTriCode  = factor(triCode,
                                 levels=mostPenaltyMinsRegSeason[["triCode"]],
                                    ordered=TRUE))  %>%
  ggplot(aes(sortedTriCode, penaltyMinutesPerGame, 
             fill=penaltyMinutesPerGame)) +
  geom_col() + 
  theme(axis.text.x=element_text(angle=90), legend.position="none") +
  scale_fill_gradient(low="blue", high="red") +
  scale_x_discrete("Team") + 
  scale_y_continuous("Penalty Minutes per Game") +
  ggtitle("Penalty Minutes per Playoff Game by Team")

plot6
```

It seems the Philly Fliers are as rowdy as Philly sports fans. I produced the same bar plot again, but this time for playoff games. I kept the ordering of most rowdy in the regular season. If there is a correlation between penalties in the regular season and in the playoffs, we will expect the order to not change much.

```{r}
plot7
```

The order did indeed not change much. Let's look at a scatter plot between penalty minutes per game in the playoffs vs. the regular season.

```{r}

plot8 <- teamTotalStats %>%
  filter(is.na(lastSeasonId)) %>%
  
  select(triCode, gameType, penaltyMinutesPerGame) %>%
  spread(gameType, penaltyMinutesPerGame) %>%
  ggplot(aes(`Regular Season`, Playoffs)) +
  geom_point(alpha=0.75,  color="blue") + 
  geom_smooth(method=lm, formula=y~x, color="red") +
  scale_x_continuous("Regular Penalty Min. per Game") +
  scale_y_continuous("Playoffs Penalty Min. per Game") + 
  ggtitle("Playoff vs. Regular Season Penalty Min. Per Game")

plot8
```

I did not expect the correlation to be that tight, but the presence of a positive correlation is not very surprising.

I wonder how penalty minutes per game relates to the win-loss ratio. Let's find out!

I created a scatter plot for the win-loss ratio vs the penalty minutes per game by game type. I experimented with the regression lines a little. I first started with a [LOESS](https://en.wikipedia.org/wiki/Local_regression) and saw an upside-down U pattern to the lines. I decided to plot them with a quadratic regression line to make things smoother.

```{r}
plot9 <- teamTotalStats %>%
  mutate(sortedTriCode  = factor(triCode,
                                 levels=mostPenaltyMinsRegSeason[["triCode"]],
                                    ordered=TRUE)) %>%
  select(sortedTriCode, gameType, penaltyMinutesPerGame, winsLossRatio) %>%
  ggplot(aes(penaltyMinutesPerGame, winsLossRatio, color=gameType)) +
  geom_point(size=2) + geom_smooth(method="lm", formula="y~poly(x, 2)") + 
  scale_x_continuous("Penalty Minutes per Game") + 
  scale_y_continuous("Win-Loss Ratio") + 
  guides(fill=guide_legend(title="Game Type")) + 
  ggtitle("Win-Loss Ratio vs. Penalty Minutes per Game by Game Type")

plot9
```

As you can see, the relationship between the win-loss ratio and penalty minutes per game is not exactly tidy, but there does seem to be a quadratic relationship there. In regular season games, it seems like 15 minutes is an optimal amount of time to be in the penalty box with regards to the win-loss ratio. For playoff games, it appears to be in the ballpark of 17 minutes.







